Chemical potential of liquids and mixtures via Adaptive Resolution Simulation

### 2014-05-11 writeDown/pub.jcp/chem-pot-jcp.tex
Luigi's revise

\bluec{
  Unified of terminology:
  \begin{itemize}
  \item interface region $\rightarrow$ transition region
  \item hybrid region $\rightarrow$ transition region
  \item define COM, change all ``center of mass'' to COM
  \item change the title of sec.III
  \item do not be too mean, we cite BOTH of their PRL H-Adress paper...
  \item some times, it is called GC-AdResS, sometimes original AdResS, sometimes standard AdResS, sometimes force-interpolation AdResS....
  \item rewrite eq. 4, see if it is OK for you
  \end{itemize}
}

\section{Introduction}
The chemical potential represents an important thermodynamic information for any system, in particular for liquids, where the possibility of combining different substances for forming optimal mixtures is strictly related to knowledge of the chemical potential of each component in the mixture environment. In this perspective, molecular simulation represents a powerful tool for predicting the chemical potential of complex molecular systems. Popular, well established methodologies in Molecular Dynamics (MD) are Widom particle insertion (IPM) \cite{widom} and thermodynamic integration (TI) \cite{ti}. IPM is computationally very demanding often beyond a reasonable limit even in presence of large computational resources, but upon convergence, is rather accurate. TI is computationally convenient but specifically designed to calculate the chemical potential and thus  it may not be optimal for employing MD for studying other properties. In fact TI requires artificial modification of the atomistic interactions (see Appendix).
Recently we have suggested that the chemical potential could be calculated by employing the Adaptive Resolution Simulation method in its Grand Canonical-like formulation (GC-AdResS) \cite{prl12,jctchan,prx}.
AdResS was originally designed  to interface regions of space at different levels of molecular resolution within one simulation set up. This allows for large and efficient multiscale simulations where the high resolution region is restricted to a small portion of space and the rest of the system is at coarser level. The recent version of the method, GC-AdResS, given its theoretical framework, should automatically calculate the chemical potential during the process of initial equilibration: in this work we prove that this is indeed the case and report results for the chemical potential for various liquids and mixtures of particular relevance in (bio)-chemistry and material science. We compare our results with those from full atomistic TI and find a satisfactory agreement. This agreement allows us to conclude that every time a multiscale GC-AdResS is performed, $\mu^{ex}$ is automatically calculated for each liquid component and implicitly confirm that the basic thermodynamics of the system is well described by the method.
\recheck{Moreover, in recent work AdResS has been merged with the MARTINI force field \cite{matej-sw1,matej-sw2}. In this context, the possibility of having a quantity like the chemical potential under \bluec{what is ``explicit control''} can be used as a further argument for the validity of the method in applications to large systems of biological interest.}
Below we provide the basic technical ingredients of GC-AdResS which are relevant for the calculation of the chemical potential, more specific details can be found in~\cite{jctchan, prx}.

@@

$w(x)$ (in green. \recheck{Here the space is devided in $x$ direction, so it is a function of only $x$.}) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\vect R}_{c}$. The extension of this additional region is equal to the cutoff radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion.\label{fig1}}

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the change of interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). {While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calcualtion of the excess chemical potential, as explain later on), the calculation of $\omega_{\thermo}$ is very relevant. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the Appendix)}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{AT}$ and $U_{CG}$ are the atomistic and coarse-grained potential. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

The result above implies that $\omega_{\thermo}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxiliary Hamiltonian  has been recently proposed and \bluec{applied to model pure and mixture systems} by Potestio {\it et al.} \cite{h-adress-0, h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{\CG}$, then  GC-AdResS can automatically provide $\mu_{\AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{\exc}_{\AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the COM is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{\AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{\dof}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calcualtion may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the \recheck{Gromacs implementation of AdResS} considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway \cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{\exc}_{\CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\begin{equation}
  \mu^{\exc}_{\AT}=
  \mu^{\exc}_{\CG}-
  \int_{\Delta}{F}_{th}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}
\label{finalmu}
\end{equation}

\section{Analytic Deriviation of the $w_{\ext}$}
{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its limitations.
We define a potential coupling the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where index $\alpha$ and $\beta$ denotes the atom position of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
%An important fact is that the force of changing representation is not
%symmetric w.r.t molecule $\alpha$ and $\beta$, therefore, the Newton's
%action-reaction law (momentum conservation) does not hold anymore for
%the potential interpolation approach. On the other hand, the force
%interpolation scheme does not preserve
%energy, while the potential interpolation
%approach has the advantage of energy conservation. In sum, both the
%force and potential interpolation approaches do not hold all physical
%conservation laws, because the hybrid region is artificially
%designed. Therefore, we will take the advantage of both and
%develop the thermodynamic relation (between the atomistic and
%coarse-grained resolutions)  for them at the same time.\\

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the transition region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume the system to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the filter). The thermodynamic force for GC-AdResS ($\vect F_\thf$) \recheck{and for the auxiliary Hamiltonian AdResS ($\vect F^\hadress_\thf$),}
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
As shown in Refs.\cite{prl12, prx}, $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $ \omega_\thf$ denotes the integral of the thermodynamic force $\vect F_\thf$. \\

Instead when we consider the auxiliary Hamiltonian approach, the third part of the R.H.S.~of Eq.~\eqref{eqn:interpol-f} is not

@@

Therefore, the change of the free energy of the system is \recheck{approximately}:

@@

\recheck{Since the system is in the thermodynamic limit, and the interactions in the system are short-ranged,
  the total free energy has the above splitting form.}
% \recheck{HAN: why is it justified to split the free energy in free energies of subsystems and why the free energy of the HYB region is only the work and not also its kinetic component. I understand that since the molecule do not change resolution, their kinetic part stays the same and thus the difference is zero. However this formula needs either a reference or few more explanations}.
$N_1$ and $N_3$ are the numbers of molecules in the AT and CG region, and
$V_1$ and $V_3$ are the volumes in the AT and CG region, respectively. $T$ is the temperature

@@

The last term comes from the work done on the piston. It is composed by
two parts, the first is the work done by the thermodynamic force, and the
second is due to 
the force of changing representation, which does not vanish due to the breaking of the Newton's
action-reaction law. The first and second part of Eq.~\eqref{eqn:interpol-f}
do not contribute, because they obey the Newton's action-reaction law, so
their total work is zero as long as the transition region move infinitesimally in one direction.
The notation $\langle\cdot\rangle$ in Eq.~\eqref{eqn:apptmp3} denotes the ensemble averge.
It is straightforward to show that \recheck{HAN, here it must be shown} \bluec{No, who cannot derive it should be sent back to the graduate course of thermodynamics.} 

@@

\recheck{The average is taken over all possible positions of the second molecule, provided the position of the first molecule
in the pairwise interaction}. In case of molecules that contain
more than one atom, the averge is also taken over all possible
conformations of the molecules. 
% By using the same argument as the Section III.C of Ref.~\cite{prx} \recheck{Han, which ones?}, we can
% show that
\bluec{here since we have the auxiliary Hamiltonian, then the system can be treated by the conclusions
  that can be found in standard thermodynamic text books. If the referee does not understand, he should
be sent back to the graduate cource. }
In the thermodynamic limit, the equilibrium volume of the AT region maxiimize the

@@

Comparing with the result of the standard GC-AdResS~Eq.~\eqref{eqn:peq},

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
Since we have an auxiliary Hamiltonian, we do not have the term $w_\ext$ in the above
fomula.
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

From Eq.~\eqref{eqn:hd-rel} and \eqref{eqn:hd-rel-2}, we
find the extra work of the thermostat in GC-AdResS being identical to the work of changing
representation of the Hamiltonian auxiliary approach:
\begin{align}\label{eqn:wextra}
  \omega_\ext = \omega_\res,
\end{align}
which basically proves the statement at the beginning of this section.  The ensemble
average on the R.H.S.~of Eq.~\eqref{eqn:w-rep} is performed in the
ensemble of the system treated with the potential interpolation approach, and the question is if the ensemble average is equivalent if it is performed in the simulation where the force
interpolation approach is used.  It is obvious that the phase spatial probability distribution correponding to the system treated with the potential interpolation is consistent with the force
interpolation at least up to the 1st order as defined in Eq.\ref{order}.  It is also
possbile to systematically obtain equivalence in the ensemble average operation  at higher orders of accuracy of the probability distribution, as, for example, it is done for the radial distribution function in Ref.~\cite{jctchan}.
However, here we do not consider higher order corrections, because it
has been numerically shown that actually the ensemble average of
$\vect F_\res$ dose not depend on in which ensemble it is
calculated~\cite{prx}. Therefore, we use Eq.~\eqref{eqn:wextra} to
calculate $\omega_\ext$, and measure the ensemble average by the
standard AdResS.  Since in the computer implementation, the CG
molecules also keep the atomistic degrees of freedom eventhough they
are in the CG region, therefore, the kinetic part of $\mu_\AT$ and
$\mu_\CG$ are identical, and $\omega_\dof$ vanishes. Therefore,
by inserting Eq.~\eqref{eqn:wextra} into \eqref{eqn:mueq-d}, we have
\begin{align}
  \mu_\AT^\exc = \mu_\CG^\exc - \int_\HY d\vect r \vect F_\thf(\vect r)
-\int_\HY d \vect r\,\nabla_{\vect r}\omega(\vect r)\,
\big\langle \omega(\vect r')\, [U^\AT(\vect r-\vect r') - U^\CG(\vect r-\vect r')]\big\rangle_{\vect r'; \vect r}
% -\int_\HY dr \nabla\omega(r)\,\big\langle \omega(r')\, [U^\AT(r-r') - U^\CG(r-r')]\big\rangle_{r'}
\end{align}
In the next section we apply the procedure to the calculation of the excess chemical potential of several liquids and mixtures.

>>>>

\section{Introduction}
The chemical potential represents an important thermodynamic information for any system, in particular for liquids, where the possibility of combining different substances for forming optimal mixtures is strictly related to knowledge of the chemical potential of each component in the mixture environment. In this perspective, molecular simulation represents a powerful tool for predicting the chemical potential of complex molecular systems. Popular, well established methodologies in Molecular Dynamics (MD) are Widom particle insertion (IPM) \cite{widom} and thermodynamic integration (TI) \cite{ti}. IPM is computationally very demanding often beyond a reasonable limit even in presence of large computational resources, but upon convergence, is rather accurate. TI is computationally convenient but specifically designed to calculate the chemical potential and thus  it may not be optimal for employing MD for studying other properties. In fact TI requires artificial modification of the atomistic interactions (see Appendix).
Recently we have suggested that the chemical potential could be calculated by employing the Adaptive Resolution Simulation method in its Grand Canonical-like formulation (GC-AdResS) \cite{prl12,jctchan,prx}.
AdResS was originally designed  to interface regions of space at different levels of molecular resolution within one simulation set up. This allows for large and efficient multiscale simulations where the high resolution region is restricted to a small portion of space and the rest of the system is at coarser level. The recent version of the method, GC-AdResS, given its theoretical framework, should automatically calculate the chemical potential during the process of initial equilibration: in this work we prove that this is indeed the case and report results for the chemical potential for various liquids and mixtures of particular relevance in (bio)-chemistry and material science. We compare our results with those from full atomistic TI and find a satisfactory agreement. This agreement allows us to conclude that every time a multiscale GC-AdResS is performed, $\mu^{ex}$ is automatically calculated for each liquid component and implicitly confirm that the basic thermodynamics of the system is well described by the method.
\recheck{Moreover, in recent work AdResS has been merged with the MARTINI force field \cite{matej-sw1,matej-sw2}. In this context, the possibility of checking the consistency of a quantity like the chemical potential can be used as a further argument for the validity of the method in applications to large systems of biological interest.}
Below we provide the basic technical ingredients of GC-AdResS which are relevant for the calculation of the chemical potential, more specific details can be found in~\cite{jctchan, prx}.

@@

$w(x)$ (in green) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\vect R}_{c}$. The extension of this additional region is equal to the cutoff radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion \recheck{(see Eq.12 of Ref.\cite{prx})}}.
\label{fig1}

@@

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the non-conservative interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calculation of the excess chemical potential, as explained later on), the calculation of $\omega_{\thermo}$ is very relevant}. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based GC-AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in GC-AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the Appendix)}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{AT}$ and $U_{CG}$ are the atomistic and coarse-grained potential.
\recheck{It must be noticed that the auxiliary Hamiltonian approach shall not be considered a Hamiltonian approach to adaptive resolution simulation. In fact, as discussed in Ref.\cite{prx} the equilibrium is imposed artificially and {\it per se} does not have any physical meaning (for more details, see discussion in the Appendix)}. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

The result above implies that $\omega_{\thermo}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxiliary Hamiltonian  has been recently proposed and applied to liquids ad mixtures (of toy models so far) by Potestio {\it et al.} \cite{h-adress-0, h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{\CG}$, then  GC-AdResS can automatically provide $\mu_{\AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{\exc}_{\AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the COM is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{\AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{\dof}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calculation may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the \recheck{Gromacs implementation of AdResS} considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway \cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{\exc}_{\CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\begin{equation}
  \mu^{\exc}_{\AT}=
  \mu^{\exc}_{\CG}-
  \int_{\Delta}{F}_{th}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}
\label{finalmu}
\end{equation}

\section{Analytic Derivation of the $w_{\ext}$}\label{sec:tmp4}
\recheck{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its conceptual limitations.
We define a potential coupling the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where index $\alpha$ and $\beta$ denotes the atom position of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the transition region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume the system to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the transition region). The thermodynamic force for GC-AdResS ($\vect F_\thf$) and for the auxiliary Hamiltonian AdResS ($\vect F^\hadress_\thf$),
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
As shown in Refs.\cite{prl12, prx}, $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $ \omega_\thf$ denotes the integral of the thermodynamic force $\vect F_\thf$. \\

Instead when we consider the auxiliary Hamiltonian approach, the third term on the R.H.S.~of Eq.~\eqref{eqn:interpol-f} is not

@@

Therefore, the change of the free energy of the system is approximately:

@@

The expression of Eq.\ref{eqn:apptmp3} as a sum of different terms is justified by the hypothesis of treating the system in the thermodynamic limit, and by the hypothesis that the interactions are short-ranged compared to the size of the transition region.
$N_1$ and $N_3$ is the numbers of molecules in the AT and CG region, and
$V_1$ and $V_3$ is the volume in the AT and CG region, respectively; $T$ is the temperature

@@

The last term is originated by the work done of the ideal piston. This latter is composed by
two parts, the first corresponding to the work done by the thermodynamic force, and the
second corresponding to 
the work done by the force of changing representation (which does not vanish due to the violation of the Newton's
action-reaction law). The first and second term of Eq.~\eqref{eqn:interpol-f}
being forces based on pairwise interactions only, do not contribute to the difference of energy; in fact 
their total work is zero (as long as the transition region move infinitesimally along $x$).
The notation $\langle\cdot\rangle$ in Eq.~\eqref{eqn:apptmp3} denotes the ensemble average.
It is straightforward to show that

@@

The average is performed over all
  possible positions of the second molecule (i.e.~$\vect r'$),
  at fixed position of the first molecule (i.e.~$\vect r$) 
  in the pairwise interaction.
In case of molecules containing
more than one atom, the average is also made over all possible
conformations in the atomistic resolution.
In the thermodynamic limit, the equilibrium volume of the AT region maximize the

@@

Comparing the expression above with that obtained for GC-AdResS~(Eq.~\eqref{eqn:peq}),

@@

In Ref.~\cite{prx} we proved that under proper assumptions,
when the flat density profile is enforced by the thermodynamic force,
the chemical potential difference between the different resolutions
is given by
\begin{align}\label{eqn:mueq-d}
 \mu_\CG - \mu_\AT =   \omega_\thf + \omega_\dof + \omega_\ext
\end{align}
The same argument can be applied to the auxiliary Hamiltonian approach,
and yields  the chemical potential difference between the AT and CG resolutions
\begin{align}\label{eqn:mueq-h}
 \mu_\CG - \mu_\AT =   \omega^\hadress_\thf + \omega_\dof 
\end{align}
In the auxiliary Hamiltonian, we do not have the term $w_\ext$ in the above
formula (being this latter, in GC-AdResS, generated by the non conservative effect of the force interpolation).
By comparing~\eqref{eqn:mueq-d} with~\eqref{eqn:mueq-h}, we have the
relation
\begin{align}\label{eqn:hd-rel-2}
  \omega_\ext = \omega^\hadress_\thf - \omega_\thf,
\end{align}
which also relates the thermodynamic force of the auxiliary Hamiltonian AdResS
and GC-AdResS.\\

From Eq.~\eqref{eqn:hd-rel} and \eqref{eqn:hd-rel-2}, we
find the extra work of the thermostat in GC-AdResS being identical to the work of changing
representation of the Hamiltonian auxiliary approach:
\begin{align}\label{eqn:wextra}
  \omega_\ext = \omega_\res,
\end{align}
which basically proves the statement at the beginning of this section.  The ensemble
average on the R.H.S.~of Eq.~\eqref{eqn:w-rep} is performed in the
ensemble of the system treated with the potential interpolation approach, and the question is if the ensemble average is equivalent if it is performed in the simulation where the force
interpolation approach is used.  It is obvious that the spatial probability distribution corresponding to the system treated with the potential interpolation is consistent with the force
interpolation at least up to the 1st order as defined in Eq.\ref{order}.  It is also
possible to systematically obtain equivalence in the ensemble average operation  at higher orders of accuracy of the probability distribution, as, for example, it is done for the radial distribution function in Ref.~\cite{jctchan}.
However, here we do not consider higher order corrections, because it
has been numerically shown that actually the ensemble average of
$\vect F_\res$ dose not depend on in which ensemble it is
calculated~\cite{prx}. Therefore, we use Eq.~\eqref{eqn:wextra} to
calculate $\omega_\ext$, and measure the ensemble average by the
standard AdResS.  As previously reported, in the computer implementation, the CG
molecules also keep the atomistic degrees of freedom even though they
are in the CG region, therefore, the kinetic part of $\mu_\AT$ and
$\mu_\CG$ are identical, and $\omega_\dof$ vanishes. Therefore,
by inserting Eq.~\eqref{eqn:wextra} into \eqref{eqn:mueq-d}, we have
\begin{align}\label{eqn:tmp24}
  \mu_\AT^\exc = \mu_\CG^\exc - \int_\HY d\vect r \vect F_\thf(\vect r)
-\int_\HY d \vect r\,\nabla_{\vect r}w(\vect r)\,
\big\langle w(\vect r')\, [\sum_{\alpha,\beta}U^\AT(\vect r_\alpha-\vect r'_\beta) - U^\CG(\vect r-\vect r')]\big\rangle_{\vect r'; \vect r}
% -\int_\HY dr \nabla\omega(r)\,\big\langle \omega(r')\, [U^\AT(r-r') - U^\CG(r-r')]\big\rangle_{r'}
\end{align}
The extension of Eq.\ref{eqn:tmp24} to multicomponent systems is reported in the Appendix, while in the next section we apply the method to the calculation of $\mu^{ex}$ to liquids and mixtures.}
